---
title: "Behavior module — validation report"
author: "Rachel Seibel"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 4
)

# Load local development tree (compile = TRUE will recompile C++ if needed)
devtools::load_all("rpops")

library(terra)
library(ggplot2)
library(tidyr)
library(dplyr)
library(yaml)

# ── Helpers ──────────────────────────────────────────────────────────────────
# Resolve extdata from local repo, fall back to installed package
local_extdata <- normalizePath(
  file.path(getwd(), "rpops", "inst", "extdata"), mustWork = FALSE
)
extdata <- if (dir.exists(local_extdata)) local_extdata else {
  file.path(system.file(package = "PoPS"), "extdata")
}
stopifnot("extdata directory not found" = dir.exists(extdata))

out_dir <- file.path(tempdir(), "behavior_validation")
dir.create(out_dir, showWarnings = FALSE)

patch_config <- function(yaml_name, subdir = "configs_pops", extra = list()) {
  src <- file.path(extdata, subdir, yaml_name)
  if (!file.exists(src)) stop("Config not found: ", src)
  cfg <- yaml::yaml.load_file(src)
  cfg$input_path  <- paste0(normalizePath(extdata, mustWork = FALSE), "/")
  cfg$output_path <- paste0(normalizePath(out_dir,  mustWork = FALSE), "/")
  cfg$random_seed <- 42L
  for (nm in names(extra)) cfg[[nm]] <- extra[[nm]]
  tmp <- tempfile(fileext = ".yaml")
  yaml::write_yaml(cfg, tmp)
  tmp
}

# Convenience: matrix → SpatRaster (unit extent, fake CRS)
mat_to_rast <- function(m) {
  r <- rast(m)
  ext(r) <- c(0, ncol(m), 0, nrow(m))
  r
}

# Pass / Fail badge
pf <- function(condition) {
  if (condition) "✅ PASS" else "❌ FAIL"
}
```

---

# Overview

This report exercises each R-layer component of the behavior module in
isolation before testing the full C++ pipeline.  Every scenario has a stated
expectation and a pass/fail verdict; figures show spatial or temporal patterns
that cannot be captured by a scalar check.

Components covered:

1. `delineate_management_units()` — unit detection and centroid placement
2. `assign_grower_types()` — distribution matching and spatial clustering
3. `grower_decision()` — detection, threshold, willingness, and `spatial_radius`
4. `behavior_configuration()` — validation of YAML-derived config lists
5. Full pipeline — spread suppression under contrasting behavioral scenarios

---

# 1  `delineate_management_units()`

## 1.1  Two-patch synthetic raster

We construct a 6 × 6 raster with two clearly separated host patches (top-left
and bottom-right 2 × 2 blocks separated by an empty column and row).
`method = "raster"` should identify exactly **2 management units**.

```{r delineate-synthetic}
# Two non-adjacent 2×2 host patches in a 6×6 grid
host_vals <- matrix(c(
  3, 3, 0, 0, 0, 0,
  3, 3, 0, 0, 0, 0,
  0, 0, 0, 0, 0, 0,
  0, 0, 0, 0, 2, 2,
  0, 0, 0, 0, 2, 2,
  0, 0, 0, 0, 0, 0
), nrow = 6, byrow = TRUE)

host_6x6 <- mat_to_rast(host_vals)

units_6x6 <- delineate_management_units(host_6x6, method = "raster")

n_units_found <- max(units_6x6$management_unit_matrix)
cat(sprintf("Units found: %d  |  Expected: 2  |  %s\n",
            n_units_found, pf(n_units_found == 2)))
print(units_6x6$management_unit_table)
```

```{r delineate-synthetic-plot, fig.height=3.5}
unit_df <- as.data.frame(mat_to_rast(units_6x6$management_unit_matrix),
                          xy = TRUE) |>
  rename(unit = lyr.1) |>
  mutate(unit = factor(ifelse(unit == 0, NA, unit)))

centroids <- units_6x6$management_unit_table

ggplot(unit_df, aes(x, y, fill = unit)) +
  geom_raster(na.rm = TRUE) +
  geom_point(data = centroids,
             aes(x = centroid_x, y = centroid_y),
             inherit.aes = FALSE, shape = 4, size = 4, stroke = 1.5) +
  scale_fill_manual(values = c("1" = "#2ecc71", "2" = "#3498db"),
                    na.value = "grey92", name = "Unit ID") +
  labs(title = "Delineated management units (synthetic 6×6)",
       subtitle = "× marks computed centroids",
       x = NULL, y = NULL) +
  coord_equal() +
  theme_minimal(base_size = 12)
```

## 1.2  Grid delineation

With `method = "grid"` and `grid_size = 3` (cells), the 6 × 6 raster is split
into four 3 × 3 blocks (rows 1–3 / 4–6 × cols 1–3 / 4–6).  Patch 1 (rows 1–2,
cols 1–2) sits entirely within the top-left block; patch 2 (rows 4–5, cols 5–6)
sits entirely within the bottom-right block.  Neither patch straddles a block
boundary, so the result should be exactly **2** non-empty units with centroids
at (1, 5) and (5, 2).

```{r delineate-grid}
# Re-use host_6x6 from section 1.1.
# grid_size = 3 places block boundaries at rows 1-3 / 4-6 and cols 1-3 / 4-6,
# so both 2×2 patches fall entirely within a single block — no splitting.
units_grid <- delineate_management_units(host_6x6, method = "grid",
                                          grid_size = 3)
# Grid IDs are global (not re-numbered to 1..N), so count distinct non-zero IDs
n_grid <- length(unique(units_grid$management_unit_matrix[
                          units_grid$management_unit_matrix > 0]))
cat(sprintf("Grid units (non-empty): %d  |  Expected: 2  |  %s\n",
            n_grid, pf(n_grid == 2)))
print(units_grid$management_unit_table)
```

---

# 2  `assign_grower_types()`

## 2.1  Deterministic edge cases

When `type_probs = c(1, 0)` every unit must receive type 1; when
`type_probs = c(0, 1)` every unit must receive type 2.

```{r assign-deterministic}
units_2x2  <- delineate_management_units(
  rast(matrix(c(1,1,1,1), nrow=2)), method = "raster"
)

# All type 1
t1 <- assign_grower_types(units_2x2$management_unit_matrix,
                            type_names = c("A","B"), type_probs = c(1,0),
                            random_seed = 1L)
all1 <- all(t1[t1 > 0] == 1L)
cat(sprintf("type_probs = c(1,0): all cells type 1?  %s\n", pf(all1)))

# All type 2
t2 <- assign_grower_types(units_2x2$management_unit_matrix,
                            type_names = c("A","B"), type_probs = c(0,1),
                            random_seed = 1L)
all2 <- all(t2[t2 > 0] == 2L)
cat(sprintf("type_probs = c(0,1): all cells type 2?  %s\n", pf(all2)))
```

## 2.2  Approximate distribution on a larger landscape

On a 10 × 10 fully-host raster the random assignment should approximately
respect the requested proportions.  We check with a tolerance of ±15 pp.

```{r assign-distribution, fig.height=3.5}
set.seed(99)

# 10×10 fully-host landscape — expect ~12 contiguous patches
host_10 <- rast(matrix(sample(c(0L, 1L), 100, replace = TRUE, prob = c(0.2, 0.8)),
                        nrow = 10))
ext(host_10) <- c(0, 10, 0, 10)

units_10 <- delineate_management_units(host_10, method = "raster")

results <- lapply(c(25, 50, 75), function(p1) {
  t_mat <- assign_grower_types(
    units_10$management_unit_matrix,
    type_names = c("early","late"),
    type_probs = c(p1, 100-p1) / 100,
    random_seed = 42L
  )
  vals <- t_mat[t_mat > 0]
  data.frame(
    target_p1 = p1,
    observed_p1 = round(mean(vals == 1) * 100, 1),
    n_cells = length(vals)
  )
}) |> bind_rows()

results$pass <- abs(results$observed_p1 - results$target_p1) <= 15
print(results)
cat(sprintf("All within ±15 pp?  %s\n", pf(all(results$pass))))

ggplot(results, aes(x = factor(target_p1), y = observed_p1)) +
  geom_col(fill = "#2980b9", width = 0.5) +
  geom_point(aes(y = target_p1), shape = 18, size = 4, colour = "#e74c3c") +
  geom_errorbar(aes(ymin = target_p1 - 15, ymax = target_p1 + 15),
                colour = "#e74c3c", width = 0.2, linetype = "dashed") +
  labs(title = "assign_grower_types(): observed vs target early-adopter %",
       subtitle = "Red diamond = target; dashed lines = ±15 pp tolerance",
       x = "Target % early adopters", y = "Observed % early adopters") +
  ylim(0, 100) +
  theme_minimal(base_size = 12)
```

## 2.3  Spatial clustering

We compare random vs clustered type assignment visually.  With a small
`cluster_range` (= 1 cell unit) the pattern should look random; with a large
`cluster_range` (= 10) spatial contiguity should be visible.

```{r assign-clustering, fig.height=4}
# Large fully-host 8×8 raster for visual clarity
host_8 <- rast(matrix(rep(1L, 64), nrow = 8))
ext(host_8) <- c(0, 8, 0, 8)
units_8 <- delineate_management_units(host_8, method = "raster",
                                        min_cells = 1)

make_type_df <- function(type_probs, spatial_structure, cluster_range = 1,
                          label) {
  mat <- assign_grower_types(
    units_8$management_unit_matrix,
    type_names = c("early","late"),
    type_probs = type_probs,
    spatial_structure = spatial_structure,
    cluster_range = cluster_range,
    random_seed = 7L
  )
  as.data.frame(mat_to_rast(mat), xy = TRUE) |>
    rename(type = lyr.1) |>
    mutate(label = label,
           type  = factor(ifelse(type == 0, NA, type),
                          levels = c(1L, 2L),
                          labels = c("early","late")))
}

df_clust <- bind_rows(
  make_type_df(c(0.5, 0.5), "random",    label = "Random"),
  make_type_df(c(0.5, 0.5), "clustered", cluster_range = 1,
               label = "Clustered (range = 1)"),
  make_type_df(c(0.5, 0.5), "clustered", cluster_range = 5,
               label = "Clustered (range = 5)")
)

ggplot(df_clust, aes(x, y, fill = type)) +
  geom_raster(na.rm = TRUE) +
  facet_wrap(~label) +
  scale_fill_manual(values = c(early = "#27ae60", late = "#e67e22"),
                    na.value = "grey90") +
  labs(title = "assign_grower_types(): spatial structure comparison",
       x = NULL, y = NULL, fill = "Grower type") +
  coord_equal() +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))
```

---

# 3  `grower_decision()`

We use a controlled 4 × 4 landscape: two management units (left two columns /
right two columns), left unit infected, right unit clean.

```{r decision-setup}
# ── Shared geometry ──────────────────────────────────────────────────────────
# 4×4 grid; unit 1 = left 2 cols, unit 2 = right 2 cols
unit_mat <- matrix(c(
  1L, 1L, 2L, 2L,
  1L, 1L, 2L, 2L,
  1L, 1L, 2L, 2L,
  1L, 1L, 2L, 2L
), nrow = 4, byrow = TRUE)

host_mat <- matrix(4L, nrow = 4, ncol = 4)   # 4 host cells everywhere

infected_mat <- matrix(c(
  2L, 2L, 0L, 0L,
  2L, 2L, 0L, 0L,
  2L, 2L, 0L, 0L,
  2L, 2L, 0L, 0L
), nrow = 4, byrow = TRUE)                   # unit 1: 50% prevalence, unit 2: 0%

# Management unit table
mut <- data.frame(unit_id   = c(1L, 2L),
                  n_cells   = c(8L, 8L),
                  centroid_x = c(1.0, 3.0),
                  centroid_y = c(2.0, 2.0))

run_decision <- function(params1, params2 = NULL, output = "continuous",
                          seed = 42L) {
  # Both units get same params unless params2 supplied
  if (is.null(params2)) params2 <- params1
  grower_type_mat <- unit_mat   # type 1 = unit 1, type 2 = unit 2
  storage.mode(grower_type_mat) <- "integer"
  grower_decision(
    infected_matrix        = infected_mat,
    host_matrix            = host_mat,
    grower_type_matrix     = grower_type_mat,
    management_unit_matrix = unit_mat,
    management_unit_table  = mut,
    behavior_params        = list(params1, params2),
    output                 = output,
    random_seed            = seed
  )
}
```

## 3.1  Always detect, always treat (detection_prob = 1, willingness = 1)

Unit 1 has 50% prevalence well above any plausible threshold.  With
`detection_prob = 1` and `willingness_to_treat = 1` it **must** be treated.
Unit 2 has 0% prevalence, so it should **not** be treated regardless.

```{r decision-always-treat}
eager <- list(detection_prob = 1.0, willingness_to_treat = 1.0,
              treatment_efficacy = 0.8, decision_threshold = 0.0)

dec_always <- run_decision(eager)
tbl_always <- attr(dec_always, "management_unit_table")

u1_treated <- tbl_always$treated[1]
u2_treated <- tbl_always$treated[2]
cat(sprintf("Unit 1 (infected) treated: %s  |  Expected TRUE  |  %s\n",
            u1_treated, pf(isTRUE(u1_treated))))
cat(sprintf("Unit 2 (clean)    treated: %s  |  Expected FALSE |  %s\n",
            u2_treated, pf(isFALSE(u2_treated))))

# Treatment map values should be 0.8 in unit 1, 0 in unit 2
u1_val <- dec_always[1, 1]
u2_val <- dec_always[1, 3]
cat(sprintf("Treatment efficacy in unit 1: %.2f  |  Expected 0.80  |  %s\n",
            u1_val, pf(abs(u1_val - 0.8) < 1e-9)))
cat(sprintf("Treatment efficacy in unit 2: %.2f  |  Expected 0.00  |  %s\n",
            u2_val, pf(abs(u2_val - 0.0) < 1e-9)))
```

## 3.2  Zero detection probability — no treatment

```{r decision-no-detect}
blind <- list(detection_prob = 0.0, willingness_to_treat = 1.0,
              treatment_efficacy = 0.8, decision_threshold = 0.0)

dec_blind <- run_decision(blind)
max_val <- max(dec_blind)
cat(sprintf("Max treatment value: %.2f  |  Expected 0.00  |  %s\n",
            max_val, pf(max_val == 0.0)))
```

## 3.3  High decision threshold blocks treatment

Unit 1 prevalence is 0.50.  Setting `decision_threshold = 0.60` means the
grower only treats above 60% prevalence — so unit 1 should remain untreated.

```{r decision-threshold}
cautious <- list(detection_prob = 1.0, willingness_to_treat = 1.0,
                 treatment_efficacy = 0.8, decision_threshold = 0.60)

dec_thresh <- run_decision(cautious)
u1_thresh <- attr(dec_thresh, "management_unit_table")$treated[1]
cat(sprintf("Unit 1 treated at threshold 0.60 (prevalence 0.50): %s  |  Expected FALSE  |  %s\n",
            u1_thresh, pf(isFALSE(u1_thresh))))
```

## 3.4  `spatial_radius` neighbourhood aggregation

We place a third small unit (unit 3) adjacent to unit 1, with a clean patch.
Unit 3's grower has `spatial_radius = 2.5`, large enough to include unit 1's
centroid (distance ≈ 1 cell).  With all-infected unit 1 in the neighbourhood,
unit 3's perceived prevalence should rise above zero even though its own patch
is clean.

```{r decision-spatial-radius}
# Extended geometry: 4×6 grid — unit 1 left, unit 2 right, unit 3 centre strip
unit_mat2 <- matrix(c(
  1L, 1L, 3L, 3L, 2L, 2L,
  1L, 1L, 3L, 3L, 2L, 2L,
  1L, 1L, 3L, 3L, 2L, 2L,
  1L, 1L, 3L, 3L, 2L, 2L
), nrow = 4, byrow = TRUE)

host_mat2     <- matrix(4L, nrow = 4, ncol = 6)
infected_mat2 <- matrix(0L, nrow = 4, ncol = 6)
infected_mat2[, 1:2] <- 3L    # unit 1: ~75% prevalence; unit 2,3: clean

mut2 <- data.frame(unit_id    = c(1L, 2L, 3L),
                   n_cells    = c(8L, 8L, 8L),
                   centroid_x = c(1.0, 5.0, 3.0),
                   centroid_y = c(2.0, 2.0, 2.0))

# Units 1,2: no spatial radius; unit 3: spatial_radius = 2.5
p_no_radius <- list(detection_prob = 1.0, willingness_to_treat = 1.0,
                    treatment_efficacy = 0.8, decision_threshold = 0.1)
p_with_radius <- modifyList(p_no_radius, list(spatial_radius = 2.5))

gtype_mat2 <- unit_mat2  # unit i → type i (1:3)
storage.mode(gtype_mat2) <- "integer"

dec_radius <- grower_decision(
  infected_matrix        = infected_mat2,
  host_matrix            = host_mat2,
  grower_type_matrix     = gtype_mat2,
  management_unit_matrix = unit_mat2,
  management_unit_table  = mut2,
  behavior_params        = list(p_no_radius, p_no_radius, p_with_radius),
  output                 = "continuous",
  random_seed            = 42L
)

tbl2 <- attr(dec_radius, "management_unit_table")
cat("Unit-level decisions with spatial_radius on unit 3:\n")
print(tbl2[, c("unit_id","infected_cells","host_cells","prevalence",
               "perceived_prevalence","treated")])

u3_pp <- tbl2$perceived_prevalence[3]
cat(sprintf("\nUnit 3 perceived prevalence: %.3f  |  Expected > 0  |  %s\n",
            u3_pp, pf(!is.na(u3_pp) && u3_pp > 0)))
```

## 3.5  Scenario comparison heatmaps

```{r decision-heatmaps, fig.height=5}
# Three contrasting parameter sets applied to the original 4×4 landscape
scenarios <- list(
  list(label = "Eager\n(det=1, wt=1, thr=0)",
       params = list(detection_prob=1.0, willingness_to_treat=1.0,
                     treatment_efficacy=0.8, decision_threshold=0.0)),
  list(label = "Threshold-limited\n(det=1, wt=1, thr=0.6)",
       params = list(detection_prob=1.0, willingness_to_treat=1.0,
                     treatment_efficacy=0.8, decision_threshold=0.6)),
  list(label = "Reluctant\n(det=0.4, wt=0.3, thr=0.1)",
       params = list(detection_prob=0.4, willingness_to_treat=0.3,
                     treatment_efficacy=0.8, decision_threshold=0.1))
)

# Run each scenario 50 times (stochastic) and take mean treatment map
n_rep <- 50
hm_df <- lapply(scenarios, function(s) {
  maps <- lapply(seq_len(n_rep), function(seed) {
    m <- run_decision(s$params, seed = seed)
    as.vector(m)
  })
  mean_map <- rowMeans(do.call(cbind, maps))

  expand.grid(row = 1:4, col = 1:4) |>
    mutate(value = mean_map,
           label = s$label)
}) |> bind_rows()

ggplot(hm_df, aes(col, rev(row), fill = value)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  facet_wrap(~label) +
  scale_fill_gradient(low = "white", high = "#c0392b",
                      limits = c(0, 0.8),
                      name = "Mean\ntreatment") +
  labs(title = "grower_decision(): mean treatment intensity across 50 replicates",
       subtitle = "Left 2 cols = infected (50% prevalence); right 2 cols = clean",
       x = "Column", y = "Row") +
  coord_equal() +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold", size = 9))
```

---

# 4  `behavior_configuration()`

```{r bc-validation}
valid_params <- list(
  list(detection_prob = 0.8, willingness_to_treat = 0.9,
       treatment_efficacy = 0.85, decision_threshold = 0.05),
  list(detection_prob = 0.4, willingness_to_treat = 0.3,
       treatment_efficacy = 0.85, decision_threshold = 0.20)
)

# ── Test 1: valid config ──────────────────────────────────────────────────────
bc_valid <- behavior_configuration(
  config_input = list(
    use_behavior_module     = TRUE,
    grower_id_file           = "",
    grower_params_file       = "",
    behavior_decision_dates  = c("2008-05-01", "2009-05-01"),
    behavior_params          = valid_params
  ),
  start_date = "2008-01-01",
  end_date   = "2009-12-31"
)
cat(sprintf("Valid config — failure field: '%s'  |  %s\n",
            ifelse(is.null(bc_valid$failure), "none", bc_valid$failure),
            pf(is.null(bc_valid$failure))))

# ── Test 2: decision date outside simulation window ───────────────────────────
bc_bad_date <- behavior_configuration(
  config_input = list(
    use_behavior_module    = TRUE,
    grower_id_file          = "",
    grower_params_file      = "",
    behavior_decision_dates = c("2010-05-01"),   # after end_date
    behavior_params         = valid_params
  ),
  start_date = "2008-01-01",
  end_date   = "2009-12-31"
)
cat(sprintf("Out-of-range date — failure: '%s'  |  %s\n",
            ifelse(is.null(bc_bad_date$failure), "none", bc_bad_date$failure),
            pf(!is.null(bc_bad_date$failure))))

# ── Test 3: defaults applied when module disabled ─────────────────────────────
bc_off <- behavior_configuration(
  config_input = list(use_behavior_module = FALSE),
  start_date = "2008-01-01", end_date = "2008-12-31"
)
cat(sprintf("Module off — decision_dates length: %d  |  %s\n",
            length(bc_off$behavior_decision_dates),
            pf(length(bc_off$behavior_decision_dates) == 0)))
```

---

# 5  Full pipeline — spread suppression scenarios

> **Requires compiled C++.**  If the package was loaded with
> `compile = FALSE` (or not yet installed from source), this section
> wraps each run in `tryCatch` and reports the compilation status.

We compare three scenarios over a 2-year monthly run on the `simple2x2`
rasters:

| Scenario | `use_behavior_module` | Grower type | Expected outcome |
|---|---|---|---|
| Baseline | FALSE | — | Unconstrained spread |
| Eager | TRUE | 100% early-adopter | Slowest spread |
| Reluctant | TRUE | 100% late-adopter | Spread near baseline |

```{r pipeline-setup}
run_pipeline <- function(label, extra_config = list()) {
  cfg_path <- patch_config(
    "simple_test_config.yaml",
    extra = c(list(number_of_iterations = 10L,
                   number_of_cores      = 1L,
                   start_date           = "2008-01-01",
                   end_date             = "2009-12-31"),
              extra_config)
  )

  config <- tryCatch(configuration(cfg_path), error = function(e) {
    message("configuration() failed: ", e$message)
    return(NULL)
  })
  if (is.null(config) || !is.null(config$failure)) {
    message("Config failure: ", config$failure)
    return(NULL)
  }

  config <- PoPS:::draw_parameters(config)
  config <- PoPS:::host_pool_setup(config)
  config$competency_table_list <-
    PoPS:::competency_table_list_creator(config$competency_table)
  config$pest_host_table_list  <-
    PoPS:::pest_host_table_list_creator(config$pest_host_table)
  config$random_seed  <- 42L
  config$random_seeds <- as.integer(config$random_seeds_list[1, ])

  result <- tryCatch(pops_model(config), error = function(e) {
    message("pops_model() failed: ", e$message)
    return(NULL)
  })
  if (is.null(result)) return(NULL)

  # Average number_infected across stochastic iterations
  inf_vec <- result$number_infected
  data.frame(
    scenario = label,
    step     = seq_along(inf_vec),
    n_infected = inf_vec
  )
}
```

```{r pipeline-baseline}
# ── 1. Baseline (no behavior) ─────────────────────────────────────────────────
df_base <- run_pipeline("Baseline")
```

```{r pipeline-behavior}
# ── 2. Eager growers ──────────────────────────────────────────────────────────
# Use the built-in behavior config (gb_base_config.yaml) as a starting point;
# override grower_id_file with a blank path so configuration() falls through
# to the behavior_params list we provide programmatically.
eager_params <- list(
  list(detection_prob = 1.0, willingness_to_treat = 1.0,
       treatment_efficacy = 0.9, decision_threshold = 0.0)
)

reluctant_params <- list(
  list(detection_prob = 0.2, willingness_to_treat = 0.2,
       treatment_efficacy = 0.9, decision_threshold = 0.4)
)

df_eager <- run_pipeline("Eager growers", extra_config = list(
  use_behavior_module     = TRUE,
  grower_id_file          = "",
  grower_params_file      = "",
  behavior_decision_dates = list("2008-06-01", "2009-06-01"),
  behavior_pesticide_duration = 30L,
  behavior_params         = eager_params
))

df_reluctant <- run_pipeline("Reluctant growers", extra_config = list(
  use_behavior_module     = TRUE,
  grower_id_file          = "",
  grower_params_file      = "",
  behavior_decision_dates = list("2008-06-01", "2009-06-01"),
  behavior_pesticide_duration = 30L,
  behavior_params         = reluctant_params
))
```

```{r pipeline-plot, fig.height=4.5}
# Collect whichever scenarios succeeded
all_dfs <- Filter(Negate(is.null), list(df_base, df_eager, df_reluctant))

if (length(all_dfs) == 0) {
  cat("⚠️  All pipeline runs failed — recompile the package and re-knit.\n")
} else {
  df_all <- bind_rows(all_dfs)

  pal <- c(Baseline         = "#7f8c8d",
           `Eager growers`  = "#27ae60",
           `Reluctant growers` = "#e67e22")

  p <- ggplot(df_all, aes(step, n_infected, colour = scenario, group = scenario)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2.5) +
    scale_colour_manual(values = pal, name = "Scenario") +
    labs(title = "Full pipeline: infected host count over time",
         subtitle = "Averaged across 10 stochastic replicates",
         x = "Output step (months)", y = "Number of infected cells") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")

  print(p)

  # Print reduction summary if all three ran
  if (length(all_dfs) == 3) {
    final <- df_all |>
      group_by(scenario) |>
      summarise(final_infected = last(n_infected), .groups = "drop")
    base_val <- final$final_infected[final$scenario == "Baseline"]
    final$reduction_pct <- round((base_val - final$final_infected) / base_val * 100, 1)
    cat("\nFinal step infection reduction vs baseline:\n")
    print(final)
    cat(sprintf(
      "Eager growers suppress more than reluctant?  %s\n",
      pf(final$reduction_pct[final$scenario == "Eager growers"] >
         final$reduction_pct[final$scenario == "Reluctant growers"])
    ))
  }
}
```

---

# 6  Summary

```{r summary-table, echo=FALSE}
summary_df <- data.frame(
  Component = c(
    "delineate_management_units()",
    "delineate_management_units() — grid",
    "assign_grower_types() deterministic",
    "assign_grower_types() distribution",
    "assign_grower_types() clustering",
    "grower_decision() always-treat",
    "grower_decision() no-detect",
    "grower_decision() threshold block",
    "grower_decision() spatial_radius",
    "behavior_configuration() valid",
    "behavior_configuration() bad date",
    "behavior_configuration() module off"
  ),
  `What is tested` = c(
    "Two separate host patches → 2 units",
    "Grid delineation produces correct unit count",
    "type_probs = c(1,0) and c(0,1) enforce exact assignment",
    "Observed proportions within ±15 pp of target",
    "cluster_range widens spatial autocorrelation",
    "det=1, wt=1: infected unit treated, clean unit not",
    "det=0: no treatment regardless of prevalence",
    "threshold > prevalence: infected unit not treated",
    "Neighbour's infection raises perceived prevalence",
    "Valid config returns no failure field",
    "Out-of-range date returns failure message",
    "Disabled module returns empty decision_dates"
  )
)
knitr::kable(summary_df, align = "ll",
             caption = "Validation checklist (detailed pass/fail in each section above)")
```

---

# Session info

```{r session-info}
sessionInfo()
```
