---
title: "England winter wheat STB — benchmark and daily vs. weekly time step comparison"
author: "Rachel Seibel"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 6
)

devtools::load_all("rpops", recompile = FALSE)

library(terra)
library(ggplot2)
library(dplyr)
library(tidyr)
library(yaml)
library(patchwork)
library(stringr)
```

---

# Overview

This notebook runs a small PoPS ensemble on the native-resolution England 2024
CROME winter wheat raster and compares a **weekly** vs. **daily** time step for
the same seed county (Lincolnshire).  A handful of stochastic iterations are run
under each setting to characterise (a) computational cost at the native grid
resolution and (b) whether the coarser weekly step materially changes spread
trajectories relative to the daily step.

---

# 1  Load England winter wheat raster

```{r load-mosaic}
# ── Adjust this path to wherever your 2024 county TIFs live ───────────────────
RASTER_DIR <- "/Users/rachelseibel/Desktop/grower_behaviour_working/rasters/2024"
if (nchar(Sys.getenv("STB_RASTER_DIR")) > 0)
  RASTER_DIR <- Sys.getenv("STB_RASTER_DIR")

tif_files <- list.files(RASTER_DIR, pattern = "_winter\\.tif$", full.names = TRUE)
cat(sprintf("Found %d county TIFs\n", length(tif_files)))

# Mosaic at native resolution — no resampling
county_rasts <- lapply(tif_files, rast)
england      <- mosaic(sprc(county_rasts), fun = "sum")
england      <- subst(england, NA, 0)   # replace NA with 0 for PoPS

cat(sprintf("Mosaic: %d rows x %d cols | res = %.0f m | CRS: %s\n",
            nrow(england), ncol(england), res(england)[1],
            crs(england, describe = TRUE)$name))
cat(sprintf("Host cells: %d | Total cells: %d\n",
            sum(values(england) > 0, na.rm = TRUE), ncell(england)))
```

```{r map-england, fig.height = 8}
# Aggregate to ~1 km for plotting only — never convert the full raster to a
# data frame (at 20 m resolution that is ~450 M rows and will exhaust memory).
PLOT_AGG <- max(1L, round(1000 / res(england)[1]))   # cells per 1 km
england_plot <- aggregate(england, fact = PLOT_AGG, fun = "sum", na.rm = TRUE)
cat(sprintf("Plot raster: %d × %d cells (aggregated %d×)\n",
            nrow(england_plot), ncol(england_plot), PLOT_AGG))

df_eng <- as.data.frame(england_plot, xy = TRUE) |>
  setNames(c("x", "y", "host")) |>
  mutate(host = ifelse(host <= 0, NA, host),
         x = x / 1000, y = y / 1000)

ggplot(df_eng, aes(x, y, fill = log1p(host))) +
  geom_raster(na.rm = TRUE) +
  scale_fill_distiller(palette = "YlGn", direction = 1,
                       name = "log(host + 1)", na.value = "grey95") +
  labs(title = "England winter wheat 2024 (CROME, native resolution)",
       subtitle = sprintf("Displayed at ~1 km aggregation | native res = %.0f m",
                          res(england)[1]),
       x = "Easting (km)", y = "Northing (km)") +
  coord_equal() +
  theme_minimal(base_size = 12)
```

---

# 2  Setup

## 2.1  Extdata paths and output directory

```{r extdata}
local_extdata <- normalizePath(file.path(getwd(), "rpops", "inst", "extdata"),
                               mustWork = FALSE)
extdata <- if (dir.exists(local_extdata)) local_extdata else
             system.file("extdata", package = "PoPS")
stopifnot("extdata not found" = dir.exists(extdata))
out_dir <- file.path(tempdir(), "stb_england_out")
dir.create(out_dir, showWarnings = FALSE)
```

## 2.2  Seed-infection raster (Lincolnshire)

```{r seed-raster}
# Approximate OSGB bounding box for Lincolnshire (metres)
LINCS_BBOX <- c(xmin = 480000, xmax = 560000, ymin = 320000, ymax = 410000)

# Crop to county, flag ~1 % of host cells as initially infected, extend back
make_seed_rast <- function(host_rast, bbox, seed_fraction = 0.01,
                            random_seed = 42L) {
  county_ext  <- ext(bbox["xmin"], bbox["xmax"], bbox["ymin"], bbox["ymax"])
  county_host <- crop(host_rast, county_ext)
  host_vals   <- as.vector(values(county_host))
  host_idx    <- which(host_vals > 0)
  if (length(host_idx) == 0) stop("No host cells in bounding box.")

  set.seed(random_seed)
  n_seed   <- max(1L, round(length(host_idx) * seed_fraction))
  seed_idx <- sample(host_idx, n_seed)

  inf_county <- county_host * 0L
  inf_vals   <- as.vector(values(inf_county))
  # Infected = 1 % of host density, minimum 1
  inf_vals[seed_idx] <- pmax(1L, as.integer(host_vals[seed_idx] * seed_fraction))
  values(inf_county) <- as.integer(inf_vals)

  # Extend to full extent, snap to host grid, replace any NA with 0
  inf_full <- extend(inf_county, host_rast, fill = 0)
  resample(inf_full, host_rast, method = "near") |> subst(NA, 0)
}

lincs_inf <- make_seed_rast(england, LINCS_BBOX)
cat(sprintf("Seed cells (Lincolnshire): %d\n",
            sum(values(lincs_inf) > 0, na.rm = TRUE)))
```

```{r seed-map, fig.height = 8}
# Aggregate for plotting at same ~1 km scale as the host map
lincs_inf_plot <- aggregate(lincs_inf, fact = PLOT_AGG, fun = "sum", na.rm = TRUE)

df_seed <- as.data.frame(lincs_inf_plot, xy = TRUE) |>
  setNames(c("x", "y", "infected")) |>
  mutate(infected = ifelse(infected <= 0, NA, infected),
         x = x / 1000, y = y / 1000)

ggplot(df_seed, aes(x, y, fill = infected)) +
  geom_raster(na.rm = TRUE) +
  scale_fill_viridis_c(option = "inferno", direction = -1,
                       name = "Infected\nunits", na.value = "grey95") +
  labs(title = "Initial infection — Lincolnshire seed",
       x = "Easting (km)", y = "Northing (km)") +
  coord_equal() +
  theme_minimal(base_size = 12)
```

## 2.3  Write rasters to temp directory

```{r write-rasters}
rast_dir <- file.path(tempdir(), "stb_england_rasts")
dir.create(rast_dir, showWarnings = FALSE)

writeRaster(england,   file.path(rast_dir, "host.tif"),     overwrite = TRUE,
            datatype = "INT4S")
writeRaster(england,   file.path(rast_dir, "total.tif"),    overwrite = TRUE,
            datatype = "INT4S")
writeRaster(lincs_inf, file.path(rast_dir, "infected.tif"), overwrite = TRUE,
            datatype = "INT4S")

# Copy required CSVs so configuration() can find them relative to input_path
csv_needed <- c("parameter_means_rr3.csv", "parameter_cov_matrix_default.csv",
                "pest_host_table_singlehost_nomort.csv",
                "competency_table_singlehost.csv",
                "prior_means_0s.csv", "prior_cov_matrix_0s.csv")
for (f in csv_needed)
  file.copy(file.path(extdata, f), file.path(rast_dir, f), overwrite = TRUE)

cat("Rasters and CSVs written to", rast_dir, "\n")
```

## 2.4  Config builder

```{r config-builder}
make_stb_config <- function(time_step = "week", n_iter = 3L) {

  # Time-step-specific parameter choices (see stb_parameterization_notes.md)
  #   Weekly: RR = 3/step, no latency, 5% long-range dispersal per step
  #   Daily:  RR = 0.5/step (~3/7), 14-day latency, 1% long-range per step
  #           (flux-equivalent: 5%/week ≈ 0.7%/day → round to 1%)
  is_daily <- time_step == "day"
  rr_csv   <- if (is_daily) "parameter_means_rr0.5.csv" else "parameter_means_rr3.csv"
  latency  <- if (is_daily) 14L else 0L
  pct_nat  <- if (is_daily) 0.99 else 0.95

  # Also copy the daily RR CSV to rast_dir if needed
  if (is_daily) {
    src <- file.path(extdata, "parameter_means_rr0.5.csv")
    file.copy(src, file.path(rast_dir, "parameter_means_rr0.5.csv"), overwrite = TRUE)
  }

  start_date <- "2024-02-01"
  end_date   <- "2024-09-30"

  cfg <- list(
    # Paths (all relative to input_path)
    input_path                    = paste0(normalizePath(rast_dir), "/"),
    output_path                   = paste0(normalizePath(out_dir),  "/"),
    host_files                    = "host.tif",
    starting_infected_files       = "infected.tif",
    total_populations_file        = "total.tif",
    exposed_files                 = "",
    parameter_means_file          = rr_csv,
    parameter_cov_matrix_file     = "parameter_cov_matrix_default.csv",
    pest_host_table_file          = "pest_host_table_singlehost_nomort.csv",
    competency_table_file         = "competency_table_singlehost.csv",
    prior_means_file              = "prior_means_0s.csv",
    prior_cov_matrix_file         = "prior_cov_matrix_0s.csv",

    # Simulation settings
    model_type                    = "SI",
    latency_period                = latency,
    start_exposed                 = FALSE,
    start_date                    = start_date,
    end_date                      = end_date,
    time_step                     = time_step,
    # Always output weekly regardless of time step — this is the key memory fix.
    # At daily resolution, 242 daily maps × raster size exhausts 16 GB.
    # Weekly output stores 35 maps instead and costs ~7× less memory.
    output_frequency              = "week",
    output_frequency_n            = 1L,
    season_month_start            = 2L,
    season_month_end              = 9L,
    number_of_iterations          = as.integer(n_iter),
    number_of_cores               = 1L,
    random_seed                   = 42L,
    use_multiple_random_seeds     = FALSE,
    county_level_infection_data   = FALSE,
    use_host_uncertainty          = FALSE,
    use_initial_condition_uncertainty = FALSE,

    # Dispersal — exponential splash + Cauchy ascospore
    # Kernel distances are in metres and do not change with time step.
    # percent_natural_dispersal is flux-adjusted: 0.95/week → 0.99/day
    # so that expected long-range events per week stay equivalent.
    natural_kernel_type           = "exponential",
    natural_distance_scale        = 300,
    natural_dir                   = "NONE",
    natural_kappa                 = 0,
    anthropogenic_kernel_type     = "cauchy",
    anthropogenic_distance_scale  = 5000,
    anthropogenic_dir             = "NONE",
    anthropogenic_kappa           = 0,
    percent_natural_dispersal     = pct_nat,

    # Stochasticity
    generate_stochasticity        = TRUE,
    establishment_stochasticity   = TRUE,
    movement_stochasticity        = TRUE,
    dispersal_stochasticity       = TRUE,
    establishment_probability     = 0.5,
    dispersal_percentage          = 0.99,

    # Weather / mortality / treatment — all off
    use_lethal_temperature        = FALSE,
    lethal_temperature_file       = "",
    lethal_temperature            = -15,
    lethal_temperature_month      = 1L,
    use_overwinter_survival       = FALSE,
    overwinter_survival_rate_month = 3L,
    overwinter_survival_rate_day  = 15L,
    overwinter_survival_rates_file = "",
    use_temperature               = FALSE,
    temperature_coefficient_file  = "",
    temperature_coefficient_sd_file = "",
    use_precipitation             = FALSE,
    precipitation_coefficient_file = "",
    weather_type                  = "deterministic",
    mortality_frequency           = "year",
    mortality_frequency_n         = 1L,
    use_treatment                 = FALSE,
    treatments_file               = "",
    treatment_dates               = start_date,
    treatment_method              = "ratio",
    pesticide_durations           = 0L,
    pesticide_efficacy            = 1.0,

    # Everything else off
    use_quarantine                = FALSE,
    quarantine_areas_file         = "",
    quarantine_directions         = "",
    use_movements                 = FALSE,
    movements_file                = "",
    use_spreadrates               = FALSE,
    use_overpopulation_movements  = FALSE,
    overpopulation_percentage     = 0.75,
    leaving_percentage            = 0.5,
    leaving_scale_coefficient     = 5,
    use_soils                     = FALSE,
    soil_starting_pest_file       = "",
    start_with_soil_populations   = FALSE,
    dispersers_to_soils_percentage = 0,
    use_mask                      = FALSE,
    mask_file                     = "",
    use_s3                        = FALSE,
    bucket                        = "",

    # Calibration off
    testing                       = FALSE,
    infected_val_cal_file         = "",
    perform_calibration           = FALSE,
    perform_validation            = FALSE,
    success_metric                = "MCC",
    calibration_method            = "ABC",
    generation_size               = 1000L,
    number_of_generations         = 7L,
    prior_number_of_observations  = 0L,
    number_of_observations        = 0L,
    params_to_estimate            = list(TRUE, TRUE, TRUE, TRUE, FALSE, FALSE),
    point_file                    = "",
    use_distance                  = FALSE,
    use_configuration             = FALSE,
    network_files                 = "",
    network_movement_types        = "walk",
    network_min_distances         = 0,
    network_max_distances         = 0,
    network_weights               = 1,
    verbose                       = FALSE
  )

  tmp <- tempfile(fileext = ".yaml")
  yaml::write_yaml(cfg, tmp)
  tmp
}
```

## 2.5  Run helper

```{r run-helper}
run_stb <- function(config_path) {
  config <- configuration(config_path)
  if (!is.null(config$failure)) stop("Config failure: ", config$failure)
  config <- PoPS:::draw_parameters(config)
  config <- PoPS:::host_pool_setup(config)
  config$competency_table_list <-
    PoPS:::competency_table_list_creator(config$competency_table)
  config$pest_host_table_list  <-
    PoPS:::pest_host_table_list_creator(config$pest_host_table)
  config$random_seed  <- 42L
  config$random_seeds <- as.integer(config$random_seeds_list[1, ])
  pops_model(config)
}
```

---

# 3  Weekly time step run

```{r run-weekly}
N_ITER <- 1L   # start with 1; increase once timing is known

cfg_weekly <- make_stb_config(time_step = "week", n_iter = N_ITER)
t_weekly   <- system.time({ res_weekly <- run_stb(cfg_weekly) })

cat(sprintf("Weekly | %d iterations | %.1f s total | %.1f s per iteration\n",
            N_ITER, t_weekly["elapsed"], t_weekly["elapsed"] / N_ITER))
cat(sprintf("Steps: %d | Final mean infected: %.0f host units\n",
            length(res_weekly$number_infected),
            res_weekly$number_infected[length(res_weekly$number_infected)]))
```

---

# 4  Daily time step run

```{r run-daily}
cfg_daily <- make_stb_config(time_step = "day", n_iter = N_ITER)
t_daily   <- system.time({ res_daily <- run_stb(cfg_daily) })

cat(sprintf("Daily  | %d iterations | %.1f s total | %.1f s per iteration\n",
            N_ITER, t_daily["elapsed"], t_daily["elapsed"] / N_ITER))
cat(sprintf("Steps: %d | Final mean infected: %.0f host units\n",
            length(res_daily$number_infected),
            res_daily$number_infected[length(res_daily$number_infected)]))
```

---

# 5  Comparison

## 5.1  Timing summary

```{r timing-table}
timing_df <- data.frame(
  `Time step`       = c("Weekly", "Daily"),
  `Steps per season`= c(length(res_weekly$number_infected),
                         length(res_daily$number_infected)),
  `Total (s)`       = round(c(t_weekly["elapsed"], t_daily["elapsed"]), 1),
  `Per iteration (s)` = round(c(t_weekly["elapsed"], t_daily["elapsed"]) / N_ITER, 1),
  check.names = FALSE
)
knitr::kable(timing_df, caption = "Computational time — Lincolnshire seed, native resolution")
```

## 5.2  Epidemic curves

Both runs use weekly output frequency, so steps are directly comparable.

```{r epidemic-curves, fig.height = 5}
epi_df <- bind_rows(
  data.frame(week      = seq_along(res_weekly$number_infected),
             n_infected = res_weekly$number_infected,
             time_step  = "Weekly (step = week)"),
  data.frame(week      = seq_along(res_daily$number_infected),
             n_infected = res_daily$number_infected,
             time_step  = "Daily (output = week)")
)

ggplot(epi_df, aes(week, n_infected, colour = time_step, group = time_step)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c("Weekly (step = week)"   = "#e74c3c",
                                  "Daily (output = week)"  = "#2980b9"),
                      name = NULL) +
  labs(title = "STB epidemic curve — weekly vs. daily time step",
       subtitle = sprintf("Lincolnshire seed | %d iteration(s) | weekly output | native resolution",
                          N_ITER),
       x = "Week of growing season", y = "Infected host units") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

---

# Session info

```{r session-info}
sessionInfo()
```
