---
title: "PoPS model — test notebook"
author: "Rachel Seibel"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
# Install rpops from github
# remotes::install_github("https://github.com/ncsu-landscape-dynamics/rpops")
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 7,
  fig.height = 4
)
devtools::load_all("rpops", compile = FALSE)
```

# Setup

```{r packages}
library(PoPS)
library(terra)
library(yaml)
library(ggplot2)
```

Locate the package's built-in test data. All configs distributed with the
package use a hard-coded Windows `input_path`; we replace it at runtime with
the correct installation path.

```{r paths}
# Prefer the local development tree: rpops/inst/extdata sits one level below
# this Rmd inside the PoPS-model-behavior repo.  When knitted from RStudio the
# working directory is the folder containing the Rmd, so the relative path
# "rpops/inst/extdata" resolves correctly.  We fall back to the installed
# package only when the local tree is absent.
local_extdata <- normalizePath(
  file.path(getwd(), "rpops", "inst", "extdata"),
  mustWork = FALSE
)

if (dir.exists(local_extdata)) {
  extdata <- local_extdata
} else {
  pkg_root <- system.file(package = "PoPS")
  if (nchar(pkg_root) == 0) {
    pkg_root <- normalizePath(file.path(getwd(), ".."), mustWork = FALSE)
  }
  extdata <- file.path(pkg_root, "extdata")
}

stopifnot("extdata directory not found — check pkg_root above" =
            dir.exists(extdata))

out_dir  <- file.path(tempdir(), "pops_test_output")
dir.create(out_dir, showWarnings = FALSE)

# Helper: load a bundled YAML, fix input_path, write to a temp file, return path
patch_config <- function(yaml_name, subdir = "configs_pops",
                          extra = list()) {
  src <- file.path(extdata, subdir, yaml_name)
  if (!file.exists(src)) stop("Config not found: ", src)
  cfg <- yaml::yaml.load_file(src)
  cfg$input_path  <- paste0(normalizePath(extdata, mustWork = FALSE), "/")
  cfg$output_path <- paste0(normalizePath(out_dir,  mustWork = FALSE), "/")
  cfg$random_seed <- 42L
  for (nm in names(extra)) cfg[[nm]] <- extra[[nm]]
  tmp <- tempfile(fileext = ".yaml")
  yaml::write_yaml(cfg, tmp)
  tmp
}
```

---

# 1  Inspect the test rasters

The `simple2x2` dataset is a 2 × 2 cell toy raster — small enough to trace by
hand and fast enough to iterate on.

```{r inspect-rasters, fig.height=3}
host_r     <- rast(file.path(extdata, "simple2x2/susceptible.tif"))
infected_r <- rast(file.path(extdata, "simple2x2/infected.tif"))
temp_r     <- rast(file.path(extdata, "simple2x2/temperature_coefficient.tif"))

par(mfrow = c(1, 3), mar = c(2, 2, 2, 1))
plot(host_r,     main = "Host density",         col = hcl.colors(10, "YlGn"))
plot(infected_r, main = "Initial infected",     col = hcl.colors(10, "Reds"))
plot(temp_r[[1]], main = "Temp coefficient [1]", col = hcl.colors(10, "Blues"))
```

Cell values and dimensions:

```{r raster-info}
data.frame(
  layer  = c("host", "infected", "temp_coeff"),
  nrow   = c(nrow(host_r), nrow(infected_r), nrow(temp_r)),
  ncol   = c(ncol(host_r), ncol(infected_r), ncol(temp_r)),
  nlyr   = c(nlyr(host_r), nlyr(infected_r), nlyr(temp_r)),
  values = c(
    paste(as.vector(values(host_r)),     collapse = ", "),
    paste(as.vector(values(infected_r)), collapse = ", "),
    paste(round(as.vector(values(temp_r[[1]])), 2), collapse = ", ")
  )
)
```

---

# 2  Minimal SI run (no weather, monthly, 1 year)

`ts6_config_t9.yaml` is the simplest single-host, no-weather, monthly config
available in the package.

```{r run-minimal}
cfg_path <- patch_config("simple_test_config.yaml",
                         extra = list(number_of_iterations = 5L,
                                      number_of_cores      = 1L))

config <- configuration(cfg_path)
stopifnot(is.null(config$failure))

config <- PoPS:::draw_parameters(config)
config <- PoPS:::host_pool_setup(config)
config$competency_table_list <- PoPS:::competency_table_list_creator(config$competency_table)
config$pest_host_table_list  <- PoPS:::pest_host_table_list_creator(config$pest_host_table)
config$random_seed  <- 42L
config$random_seeds <- as.integer(config$random_seeds_list[1, ])

result <- pops_model(config)
```

## 2.1  Infected counts over time

```{r plot-infected-counts}
infected_df <- data.frame(
  step            = seq_along(result$number_infected),
  number_infected = result$number_infected
)

ggplot(infected_df, aes(step, number_infected)) +
  geom_line(color = "#c0392b", linewidth = 0.8) +
  geom_point(color = "#c0392b", size = 2) +
  labs(title = "Infected host count over time",
       x = "Output step", y = "Number of infected cells") +
  theme_minimal(base_size = 12)
```

## 2.2  Final infected raster

```{r plot-final-raster, fig.height=3}
last_step   <- length(result$host_pools[[1]]$infected)
final_inf   <- rast(result$host_pools[[1]]$infected[[last_step]])
ext(final_inf) <- ext(host_r)
crs(final_inf) <- crs(host_r)

par(mfrow = c(1, 2), mar = c(2, 2, 2, 1))
plot(infected_r,  main = "t = 0",   col = hcl.colors(10, "Reds"), range = c(0, 3))
plot(final_inf,   main = "t = end", col = hcl.colors(10, "Reds"), range = c(0, 3))
```

## 2.3  Output list structure

```{r output-structure}
str(result, max.level = 2)
```

---

# 3  Multi-year run with weather (monthly, 3 years)

`ts9_config_t5.yaml` adds a temperature coefficient raster — the model
multiplies the reproductive rate by the weather coefficient at each step.

```{r run-weather}
cfg_path_wx <- patch_config("simple_test_config.yaml",
                            extra = list(number_of_iterations = 5L,
                                         number_of_cores      = 1L,
                                         start_date           = "2008-01-01",
                                         end_date             = "2010-12-31"))

config_wx <- configuration(cfg_path_wx)
stopifnot(is.null(config_wx$failure))

config_wx <- PoPS:::draw_parameters(config_wx)
config_wx <- PoPS:::host_pool_setup(config_wx)
config_wx$competency_table_list <- PoPS:::competency_table_list_creator(config_wx$competency_table)
config_wx$pest_host_table_list  <- PoPS:::pest_host_table_list_creator(config_wx$pest_host_table)
config_wx$random_seed  <- 42L
config_wx$random_seeds <- as.integer(config_wx$random_seeds_list[1, ])

result_wx <- pops_model(config_wx)

data.frame(
  step           = seq_along(result_wx$number_infected),
  n_infected     = result_wx$number_infected,
  area_infected  = result_wx$area_infected
)
```

```{r plot-weather-run}
ggplot(data.frame(step = seq_along(result_wx$number_infected),
                  n    = result_wx$number_infected),
       aes(step, n)) +
  geom_line(color = "#2980b9", linewidth = 0.8) +
  geom_point(color = "#2980b9", size = 2) +
  labs(title = "Infected count — 3-year run with temperature coefficient",
       x = "Annual output step", y = "Number of infected cells") +
  theme_minimal(base_size = 12)
```

---

# 4  Behavior module pre-processing

The C++ behavior module is not yet wired into `pops_model_cpp()`.  These
chunks test the R pre-processing functions that will supply it with inputs.

## 4.1  Delineate management units

```{r delineate}
units <- delineate_management_units(host_r, method = "raster")

cat("Unit IDs found:", sort(unique(as.integer(
  units$management_unit_matrix[units$management_unit_matrix > 0]
))), "\n")

print(units$management_unit_table)
```

```{r plot-units, fig.height=3}
unit_rast <- rast(units$management_unit_matrix)
ext(unit_rast) <- ext(host_r)
crs(unit_rast) <- crs(host_r)
plot(unit_rast, main = "Management unit IDs",
     col = hcl.colors(max(units$management_unit_matrix), "Set2"))
```

## 4.2  Assign grower types

Assign two behavioral types (early adopter, late adopter) with equal prior
probability:

```{r assign-types}
type_mat <- assign_grower_types(
  management_units  = units$management_unit_matrix,
  type_names        = c("early_adopter", "late_adopter"),
  type_probs        = c(0.5, 0.5),
  spatial_structure = "random",
  random_seed       = 42L
)

cat("Type assignment (1 = early adopter, 2 = late adopter):\n")
print(type_mat)
```

## 4.3  Grower decision at a single time step

```{r grower-decision}
infected_mat <- as.matrix(infected_r, wide = TRUE)
storage.mode(infected_mat) <- "integer"
host_mat     <- as.matrix(host_r,     wide = TRUE)
storage.mode(host_mat)     <- "integer"

behavior_params <- list(
  list(detection_prob       = 0.8,   # early adopter: observant
       willingness_to_treat = 0.9,
       treatment_efficacy   = 0.85,
       decision_threshold   = 0.05),
  list(detection_prob       = 0.4,   # late adopter: less responsive
       willingness_to_treat = 0.3,
       treatment_efficacy   = 0.85,
       decision_threshold   = 0.20)
)

decision <- grower_decision(
  infected_matrix        = infected_mat,
  host_matrix            = host_mat,
  grower_type_matrix     = type_mat,
  management_unit_matrix = units$management_unit_matrix,
  management_unit_table  = units$management_unit_table,
  behavior_params       = behavior_params,
  output                 = "continuous",
  random_seed            = 42L
)

cat("Treatment map (efficacy values):\n")
print(decision)

tbl <- attr(decision, "management_unit_table")
cat("\nUnit-level infection summary:\n")
print(tbl[, c("unit_id", "infected_cells", "host_cells", "prevalence")])
```

## 4.4  behavior_configuration()

Validate a configuration list with the new field set:

```{r bc-validate}
bc <- behavior_configuration(
  config_input = list(
    use_behavior_module     = TRUE,
    grower_id_file           = file.path(extdata, "simple2x2/host.tif"),
    grower_params_file       = "",
    behavior_decision_dates = c("2008-05-01", "2008-09-01"),
    behavior_params         = behavior_params
  ),
  start_date = "2008-01-01",
  end_date   = "2008-12-31"
)

cat("Failure:", ifelse(is.null(bc$failure), "none", bc$failure), "\n")
cat("Module enabled:", bc$use_behavior_module, "\n")
cat("Decision dates:", paste(bc$behavior_decision_dates, collapse = ", "), "\n")
cat("Type 1 detection_prob:", bc$behavior_params[[1]]$detection_prob, "\n")
cat("Type 2 decision_threshold:", bc$behavior_params[[2]]$decision_threshold, "\n")
```

---

# 5  Session info

```{r session-info}
sessionInfo()
```
